---
title: "Chapter 17"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE, fig.align = "center")
library(tidyverse)
library(janitor)
```

# Confidence Intervals for Means

**Objectives:**

________________

I.    Discover the sampling distribution of $\bar{x}$

II.   Use the distribution to solve problems

III.  $t$ distribution 

IV.   Examples of CI for $\mu$

V.    Special conditions for $t$

________________

```{r, echo = FALSE}
opar <- par(no.readonly = TRUE)
# set.seed(33)
par(mfrow=c(4, 3))
par(mar= c(2, 2, 3, 1))
m <- 100000  # Number of Samples
EX <- 1.1   # Character expansion
COLOR = "lightblue"
# X~N(5,5)
x <- seq(-12, 22,.01)
y <- dnorm(x,5,5)
plot(x, y, type = "l",col = "black", lwd = 2, main = "X ~ N(5, 5)",xlab="", ylab = "")
# X~U((10 - sqrt(300))/2, (10 + sqrt(300))/2))
x <- seq(-12, 22,.01)
y <- dunif(x,(10 - sqrt(300))/2, (10 + sqrt(300))/2)
plot(x, y, type = "l", col = "black", lwd = 2, main = "X ~ U(-3.66, 13.66)", xlab = "", ylab = "")
# X~Exp(1/5)
x <- seq(-12, 22,.01)
y <- dexp(x, 1/5)
plot(x, y, type = "l", col = "black", lwd = 2, main = "X ~ Exp(1/5)", xlab = "", ylab = "")
xbar.2 <- apply(matrix(rnorm(m*2,5,5),nrow=m),1,mean)
n2<- c(mean(xbar.2), sd(xbar.2))
hist(xbar.2,breaks=30,col = COLOR, xlim=c(-12, 22),prob=T,xlab="",ylab="",main="")
mtext(expression(bar(x)[2]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(2)), add = TRUE, n = 300, lwd = 2)
xbar.2 <- apply(matrix(runif(m*2,(10 - sqrt(300))/2, (10 + sqrt(300))/2), nrow=m),1,mean)
u2 <- c(mean(xbar.2), sd(xbar.2))
hist(xbar.2,breaks=50,col = COLOR,xlim=c(-12, 22),prob=T,xlab="",ylab="",main= "")
mtext(expression(bar(x)[2]), side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(2)), add = TRUE, n = 300, lwd = 2)
xbar.2 <- apply(matrix(rexp(m*2,1/5),nrow=m),1,mean)
e2 <- c(mean(xbar.2), sd(xbar.2))
hist(xbar.2,breaks=50,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[2]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(2)), add = TRUE, n = 300, lwd = 2)
xbar.12 <- apply(matrix(rnorm(m*12,5,5),nrow=m),1,mean)
n12 <- c(mean(xbar.12), sd(xbar.12))
hist(xbar.12,breaks=30,col = COLOR,xlim=c(-12, 22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[12]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(12)), add = TRUE, n = 300, lwd = 2)
xbar.12 <- apply(matrix(runif(m*12,(10 - sqrt(300))/2, (10 + sqrt(300))/2), nrow=m),1,mean)
u12 <- c(mean(xbar.12), sd(xbar.12))
hist(xbar.12,breaks=30,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[12]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(12)), add = TRUE, n = 300, lwd = 2)
xbar.12 <- apply(matrix(rexp(m*12,1/5),nrow=m),1,mean)
e12 <- c(mean(xbar.12), sd(xbar.12))
hist(xbar.12,breaks=30,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[12]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(12)), add = TRUE, n = 300, lwd = 2)
xbar.30 <- apply(matrix(rnorm(m*30,5,5),nrow=m),1,mean)
n30 <- c(mean(xbar.30), sd(xbar.30))
hist(xbar.30,breaks=20,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[30]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(30)), add = TRUE, n = 300, lwd = 2)
xbar.30 <- apply(matrix(runif(m*30,(10 - sqrt(300))/2, (10 + sqrt(300))/2), nrow=m),1,mean)
u30 <- c(mean(xbar.30), sd(xbar.30))
hist(xbar.30,breaks=20,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[30]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(30)), add = TRUE, n = 300, lwd = 2)
xbar.30 <- apply(matrix(rexp(m*30,1/5),nrow=m),1,mean)
e30 <- c(mean(xbar.30), sd(xbar.30))
hist(xbar.30,breaks=20,col = COLOR,xlim=c(-12,22),prob=T,xlab="",ylab="",main="")
mtext(  expression(bar(x)[30]),side=3,line=1,cex=EX)
curve(dnorm(x, 5, 5/sqrt(30)), add = TRUE, n = 300, lwd = 2)
par(opar)
matrix(c(n2, u2, e2, n12, u12, e12, n30, u30, e30), byrow = TRUE, nrow = 3)
```

```{r}
m <- 100000  # Number of Samples
EX <- 1.1   # Character expansion
COLOR = "lightblue"
xbar.30 <- apply(matrix(rexp(m*30,1/5),nrow=m),1,mean)
e30 <- c(mean(xbar.30), sd(xbar.30))
hist(xbar.30, breaks = 70, col = COLOR, xlim=c(1, 9), prob=TRUE, xlab="", ylab="", main="")
mtext(expression(bar(x)[30]), side = 3, line = 1,cex = EX)
curve(dnorm(x, 5, 5/sqrt(30)), add = TRUE, n = 300, lwd = 2)
```
If $n$ is "large enough" and sampled values are independent, $\bar{x} \dot\sim \mathcal{N}\left(\mu_{\bar{x}} = \mu, \sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}} \right)$

**Conditions:**

1. **Randomization**---data values are sampled randomly OR sample is representative
2. **10%**---Less than 10% of the population is sampled.
3. **Large enough sample**---hsape of the distribution looks normal.  Sampling from a normal...any size; sampling from a symmetric distribution...12 may be sufficient; sampling from a very skewed population....may need 1500 or more.

> **The Central Limit Theorem** states that as $n$ increases, the sample mean has a more and more normal shaped sampling distribution with the same mean as the population and a standard deviation of $\sigma/\sqrt{n}$.

________

## Example

Wildlife scientists studying a certain species of frogs know that adults should weigh an average of 118 grams with a standard deviation of 14 grams.

1. What is the distribution of $\bar{x}$?  

**ANSWER:** $\bar{x} \sim \mathcal{N}(118, 14/\sqrt{50} = `r 14/sqrt(50)`)$

2. A random sample of 50 adult frogs gives a mean weight of only 110 grams,  Is this unusually low?  

**ANSWER:** 

3.  What percent of similar samples would have average weights less than 112 grams?

$P(\bar{x}\leq 112)=$?

```{r}
# Your code here
```

4. Between 115 and 120 grams?

$P(115 \leq \bar{x} \leq 120)=$?

```{r}
# Your code here
```

5.  More than 125 grams?

$P(\bar{x} \geq 125)=$?

```{r}
# Your code here
```

6. What is the 5$^{\text{th}}$ percentile of average from weight when $n = 50$?

$P(\bar{x} \leq q) = 0.05\rightarrow q = `r qnorm(0.05, 118, 14/sqrt(50))`$?

```{r}
# Your code here
```
________________

Problem with the normal distribution is that we seldom know the value of $\sigma$!  

What happens when $\sigma$ is replaced with $s$ in $\frac{(\bar{X} - \mu)}{\sigma/\sqrt{n}}$?

$$\frac{(\bar{X} - \mu)}{s/\sqrt{n}} \sim t_{n-1}$$

$$f(t) = \frac{\Gamma(\frac{\nu + 1}{2})}{\sqrt{\nu\pi}\cdot\Gamma(\frac{\nu}{2})}\left(1 + \frac{t^2}{\nu} \right)^{-\tfrac{\nu+1}{2}}$$
where $\nu =$ df, $\Gamma(n) = (n - 1)!$, $\Gamma(\frac{1}{2}) = \sqrt{\pi}$, and $\Gamma(x)=\int_{0}^{\infty}t^{x -1}e^{-t}dt$

__________

**To use a $t-$distribution, sample size determines what to check**

$n < 20$ check that sample is enarly normal; $20 \leq n < 30$ sample should have no outliers, some skewness OK; $n \geq 30$ sample should not have strong skewnss or far outliers.

____________

## Example

A coffee machine dispenses coffee into paper cups.  The machine says it dispenses 10 oz, but the actual amount varies from cup to cup.  Actual amount for the 20 cups are: 9.9, 10.0, 9.9, 10.0, 9.7, 9.6, 9.5, 9.6, 9.9, 10.0, 9.8, 9.7, 10.2, 9.5, 10.1, 9.8, 10.1, 9.8, and 9.9.

Is there evidence the machine is shortchanging customers per a confidence interval?  If so, by how much?

```{r}
# read in data
coffee <- c(9.9, 10.0, 9.9, 10.0, 9.7, 9.6, 
            9.5, 9.6, 9.9, 10.0, 9.8, 9.7, 
            10.2, 9.5, 10.1, 9.8, 10.1, 9.8, 9.9)
DF <- data.frame(coffee = coffee)
# assess data
# using just a vector
qqnorm(coffee)
qqline(coffee)
# note: argument is not x but sample here
ggplot(data = DF, aes(sample = coffee)) +
  stat_qq() + 
  stat_qq_line() + 
  theme_bw()
# construct CI
t.test(x = coffee, conf.level = 0.95)
# Or
t.test(x = coffee, conf.level = 0.95)$conf
# Or using formula
n <- sum(!is.na(coffee))
xbar <- mean(coffee)
S <- sd(coffee)
CT <- qt(0.975, n-1)
ME <- CT*S/sqrt(n)
LL <- xbar - ME
UL <- xbar + ME
c(xbar, S, CT, ME, LL, UL)
```

__________

## Example

A company has set a goal of developing a battery that lasts over 5 hours (300 minutes) of continuous use.  The first 12 batteries lasted: 321, 295, 332, 351, 281, 336, 311, 253, 270, 326, 311, and 288 minutes. Is there evidence the company has met its goal per a confidence interval?

```{r}
battery <- c(321, 295, 332, 351, 281, 336, 
             311, 253, 270, 326, 311, 288)
DF2 <- data.frame(battery = battery)
# Your finish the code

```
_________

## Bootstrap percentile CI

* Take $B$ samples of the same size ($n$) as the original sample with replacement from the original sample.  These $B$ samples are called bootstrap samples.
* Compute a statistic of interest (sample mean for the two previous examples) for each bootstrap sample.
* Find the quantiles (say 0.05 and 0.95 for a 90% CI) of the bootstrap distribution.  

```{r}
B <- 10000
xbar <- numeric(B)
for(i in 1:B){
  bss <- sample(coffee, size = 20, replace = TRUE)
  xbar[i] <- mean(bss)
}
BPCI <- quantile(xbar, probs = c(0.025, 0.975))
BPCI
#
xbar <- numeric(B)
for(i in 1:B){
  bss <- sample(battery, size = 12, replace = TRUE)
  xbar[i] <- mean(bss)
}
BPCI <- quantile(xbar, probs = c(0.025, 0.975))
BPCI
```